---
title: "Machine Learning - Cource Project"
output:
  html_document: default
  pdf_document: default
date: "2022-08-24"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Hagit Glickman

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

## Data
The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. 

The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

## Goal
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 


## Loading Libraries

```{r}

library(lattice)
library (ggplot2)
library(caret)
library(dplyr)
library(groupdata2)
library(randomForest)
```
## Reading Data

pml_training contains 19622 observations and 160 variables.

pml_testing contains 20 observations and 160 variables.

```{r}
setwd("C:/Users/Administrator/Documents/My R/Machine Learning/project")
pml_training <- read.csv("pml_training.csv")
pml_testing <- read.csv("pml_testing.csv")
```

## Cleaning data

The data include computed summery variables like max, min, avg, kurtosis and etc..
The summery variables are computed only at the end of activity windows and hence include 
large amount of missing values. In order to remove these variables a function 'drop_from_df' is defined.
It gets a data file and a string as an input and return the data file without any variable (column) with a name containing the string.  


```{r}
drop_from_df <- function(df, string) {
  string <- df[ , grepl(string,colnames(df))]
  string_name <- names(string)
  df <- df[ , !(names(df) %in% string_name)]
  return(df)
}
```

The computed summery variables are removed from both data files -training and testing.
Variables like X, user_name and new_window are also removed from the analysis. 

After the cleaning:

pml_training contains 19622 observations and 53 variables.

pml_testing contains 20 observations and 53 variables.


```{r}
pml_training <- drop_from_df(pml_training, "kurtosis")
pml_training <- drop_from_df(pml_training, "skewness") 
pml_training <- drop_from_df(pml_training, "max")
pml_training <- drop_from_df(pml_training, "min")
pml_training <- drop_from_df(pml_training, "amplitude")
pml_training <- drop_from_df(pml_training, "avg")
pml_training <- drop_from_df(pml_training, "stddev")
pml_training <- drop_from_df(pml_training, "var")
pml_training <- subset(pml_training, select=-c(X, user_name, raw_timestamp_part_1,
                                               raw_timestamp_part_2, cvtd_timestamp,
                                               new_window, num_window))


pml_testing <- drop_from_df(pml_testing, "kurtosis")
pml_testing <- drop_from_df(pml_testing, "skewness") 
pml_testing <- drop_from_df(pml_testing, "max")
pml_testing <- drop_from_df(pml_testing, "min")
pml_testing <- drop_from_df(pml_testing, "amplitude")
pml_testing <- drop_from_df(pml_testing, "avg")
pml_testing <- drop_from_df(pml_testing, "stddev")
pml_testing <- drop_from_df(pml_testing, "var")
pml_testing <- subset(pml_testing, select=-c(X, user_name, raw_timestamp_part_1,
                                               raw_timestamp_part_2, cvtd_timestamp,
                                               new_window, num_window))
```
## Spliting data and validation

In order to evaluate and choose an appropriate classification model,
the training data is split into two sets - training (70%) and validation (30%). 
The training data is used to build 3 different classification models, LDA, Naive
Bayes and Random Forest. The validation data is used to evaluate the models and choose 
the model with the highest accuracy level. 


After cleaning and splitting there are 3 working data sets:

training - 13735 observations and 53 variables.
validation - 5887 observations and 53 variables.
pml_testing - 20 observations and 53 variables.


```{r}
set.seed (9876)

pml_training$classe = factor(pml_training$classe)
inTrain <- createDataPartition(y=pml_training$classe, p = 0.7, list=FALSE)
training <- pml_training[inTrain,]
validation <- pml_training[-inTrain,]

```

## Ploting predictors

Before starting to build a classification modal, a series of plots were generated 
in order to understand how the data actually look and how the data interact with 
each other. The following plot is an example, showing the distribution of 3 different 
measures 'roll', 'pitch' and 'yaw', measured on 4 different accelerometers located on 
the belt, forearm, arm  and dumbell, as a function of the 5 ways of performing barbell 
lifts A-E - the 'classe' variable. 

As we can learn from the plot there are differences between the distributions of predictors
across ways of barbell lifts. These differences are useful in classifying problem.
It is also clear that it will be better to standardize the predictors as the scales
changes quite a lot ((-50,50) to (-150,150)).

```{r}
par(mfrow=c(3,4))

boxplot(roll_belt~classe,
        data=pml_training,
        main="'roll_belt'",
        xlab="",
        ylab="",
        col="skyblue2",
        border="black"
)
boxplot(roll_arm~classe,
        data=pml_training,
        main="'roll_arm'",
        xlab="",
        ylab="",
        col="darksalmon",
        border="black"
)
boxplot(roll_dumbbell~classe,
        data=pml_training,
        main="'roll_dumbbell'",
        xlab="",
        ylab="",
        col="mediumaquamarine",
        border="black"
)
boxplot(roll_forearm~classe,
        data=pml_training,
        main="'roll_forearm'",
        xlab="",
        ylab="",
        col="plum2",
        border="black"
)
boxplot(pitch_belt~classe,
        data=pml_training,
        main="'pitch_belt'",
        xlab="",
        ylab="",
        col="skyblue2",
        border="black"
)
boxplot(pitch_arm~classe,
        data=pml_training,
        main="'pitch_arm'",
        xlab="",
        ylab="",
        col="darksalmon",
        border="black"
)
boxplot(pitch_dumbbell~classe,
        data=pml_training,
        main="'pitch_dumbbell'",
        xlab="",
        ylab="",
        col="mediumaquamarine",
        border="black"
)
boxplot(pitch_forearm~classe,
        data=pml_training,
        main="'pitch_forearm'",
        xlab="",
        ylab="",
        col="plum2",
        border="black"
)
boxplot(yaw_belt~classe,
        data=pml_training,
        main="'yaw_belt'",
        xlab="way of barbell lift",
        ylab="",
        col="skyblue2",
        border="black"
)
boxplot(yaw_arm~classe,
        data=pml_training,
        main="'yaw_arm'",
        xlab="way of barbell lift",
        ylab="",
        col="darksalmon",
        border="black"
)
boxplot(yaw_dumbbell~classe,
        data=pml_training,
        main="'yaw_dumbbell'",
        xlab="way of barbell lift",
        ylab="",
        col="mediumaquamarine",
        border="black"
)
boxplot(yaw_forearm~classe,
        data=pml_training,
        main="'yaw_forearm'",
        xlab="way of barbell lift",
        ylab="",
        col="plum2",
        border="black"
)
```

## Centering and Scaling

```{r}
preObj <- preProcess(training[,-53], method = c("center", "scale"))

training_s <- predict(preObj, training[,-53])
classe <-  training$classe
training_s <- cbind (training_s, classe)

validation_s <- predict(preObj, validation[,-53])
classe <-  validation$classe
validation_s <- cbind (validation_s, classe)

testing_s <-predict(preObj, pml_testing[,-53])
testing_s <- cbind (testing_s, pml_testing$problem_id)
```

##  The LDA Classifier
```{r}
repeat_cv <- trainControl(method='repeatedcv', number=5, repeats=3)
classifier_lda <- train(classe~., 
                 data=training_s,
                 method="lda",
                 trControl=repeat_cv,
                 metric='Accuracy',
                 prox=TRUE)
classifier_lda
prediction_lda <- predict(classifier_lda, newdata=validation_s)
confusionMatrix(prediction_lda, validation_s$classe)
```


## The Naive Bayes Classifier

```{r}
repeat_cv <- trainControl(method='repeatedcv', number=5, repeats=3)
classifier_NB <- train(classe~., 
                       data=training_s,
                       method="nb",
                       trControl=repeat_cv,
                       metric='Accuracy',
                       prox=TRUE)
classifier_NB
prediction_NB <- predict(classifier_NB, newdata=validation_s) 
confusionMatrix(prediction_NB, validation_s$classe)
```

## The Random Forest Classifier

```{r}
classifier_RF = randomForest(classe~., data=training_s, ntree = 500)
classifier_RF
prediction_RF <- predict(classifier_RF, newdata=validation_s) 
confusionMatrix(prediction_RF, validation_s$classe)

```

The Random Forests gave an Accuracy of 99.42%, which is very high and higher than the LDA (70.35%)
and the Naive Bayes (74.55%). 

The final model is Random Forest. The expected out-of-sample error is 100-99.42 = 0.58%.

## Predicting Test Cases

```{r}
prediction_RF_test <- predict(classifier_RF, newdata=testing_s)
prediction_RF_test 
```

## END









